---
title: "DRAFT: ESA compliance with remotely sensed data"
author: "Tiffany Kim, Ya-Wei Li, and Jacob Malcom; Defenders of Wildlife"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    code_folding: hide
    css: custom.css
    df_print: paged
    fig_caption: yes
    fig_width: 7.5
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
---

## Overview

### Section 7 compliance monitoring analyses

This vignette captures the analyses in a forthcoming manuscript (preprint + 
peer-reviewed paper) evaluating the utility of remotely sensed data for monitoring
compliance with section 7 of the US Endangered Species Act. The section 7 
consultation data - work types, coordinates, etc. - are from the TAILS database
from the US Fish and Wildlife Service (FWS); see 
[Malcom and Li 2015, _PNAS_](http://www.pnas.org/content/112/52/15844.abstract) 
for more information.

```{r setup, include=FALSE}
library(dplyr)
library(ggalt)
library(ggmap)
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(highcharter)
library(knitr)
library(leaflet)
library(lubridate)
library(plotly)
library(readr)
library(readxl)
library(stringr)
library(superheat)
library(tidyr)
library(viridis)

inf <- system.file("extdata", "merged_formal_informal.rds", 
                   package = "compliance.exploration")
dat <- readRDS(inf)
dat$earliest_date <- as.Date(dat$earliest_date)
n_obs <- dim(dat)[1]
formal <- filter(dat, formal_in == "formal")
informal <- filter(dat, formal_in == "informal")

form_rand <- system.file("extdata", "formal_consult_random.rds",
                         package = "compliance.exploration")
inform_rand <- system.file("extdata", "informal_consult_random.rds",
                           package = "compliance.exploration")
form_cons <- readRDS(form_rand)
inform_cons <- readRDS(inform_rand)

all_s7 <- system.file("extdata", "s7_data.rds",
                      package = "compliance.exploration")
s7_all <- readRDS(all_s7)
```

## Data Summary

The dataset includes `r n_obs` consultations randomly selected from among 
over 44,000 consultations recorded with coordinates by FWS from 2008-2015.

### Top of the data

```{r datahead}
head(dat)
dim(dat)
```

### Formal and informal consultation

Although FWS's consultation data shows that [formal consultations](https://www.fws.gov/endangered/what-we-do/faq.html) only make up
~7% of all consultations, actions only undergo formal consultation if an ESA-
listed species is likely to be harmed. That is, we are generally more interested
in actions that have undergone formal consultation. The 182 formal consultations
we investigated are therefore a "denser" sampling of all formal consultations 
than the 182 samples of informal consultations.

### Geography

The locations of the consultations in our sample generally reflected the 
distribution of consultations observed by [Malcom and Li (2015)](http://www.pnas.org/content/112/52/15844.abstract):
formal consultations were more likely to come from the western US, informal 
consultations more likely from the East. The bias in which offices record 
coordinates is apparent with the lack of formal consultations from some states,
like Colorado (n ~ 500 formal consultations, but rarely records coordinates) and
Washington (n ~ 500 formal consultations, high data recording rate).

```{r map}
### A static map in case we want one for Supp. Info.
#
# to_map <- dplyr::filter(dat, dat$lat_dec_deg.x > 20 & dat$lat_dec_deg.x < 60)
# map <- qmplot(
#     data = to_map,
#     x = long_dec_deg.x,
#     y = lat_dec_deg.x,
#     maptype = "toner-lite",
#     color = formal_in,
#     size = I(2),
#     alpha = I(0.5),
#     extent = "normal",
#     zoom = 5) +
#   theme_bw() +
#   labs(x = "Longitude",
#        y = "Latitude") +
#   scale_color_viridis(discrete = TRUE, name = "Type")
# ggsave(plot = map, "inst/figs/map.pdf")

cols <- substr(
  viridis(3), 0, 7
)

dat$formal_colors <- ifelse(
  dat$formal_in == "formal",
  cols[1],
  cols[3]
)

tags$div(
  style = "background-color: #e9e9e9",
  leaflet(dat) %>%
    setView(lng=-95, lat=38, zoom = 4) %>%
    addProviderTiles("Stamen.TonerLite") %>%
    mapOptions(zoomToLimits = "first") %>%
    addCircleMarkers(
      lng= ~long_dec_deg.x, 
      lat= ~lat_dec_deg.x,
      stroke = FALSE,
      fillOpacity = 0.5,
      fillColor = ~formal_colors,
      popup = ~activity_code
    )
)
```

### Other variables summaries

__Top lead agencies__

```{r agency} 
as_data_frame(head(sort(table(as.character(dat$lead_agency)), decreasing = TRUE), 10))
```

__Top 10 lead ES offices__

```{r ESO} 
head(sort(table(as.character(dat$ESOffice.x)), decreasing = TRUE), 10)
```

__Fiscal years__

```{r FY} 
table(dat$FY)
```

## Action Observability

### Majority of actions not found

Before starting to explore the remotely sensed data, two authors (TK and JM) 
independently scored each action type for _expected_ observability, with:

- 0: not expected to be observered;
- 0.5: may be observed or unknown whether to expect; and
- 1: expected to be observed.

After independent scoring, we met and reconciled the expected scores. The authors 
- primarily TK - then attempted to find all `r n_obs` consultations in aerial
images using Google Earth and the coordinates provided by FWS. 

First, the overall expected and observed observability for formal and informal 
consultations:

```{r p1}
expected <- aggregate(reconcile ~ formal_in, 
                      data = dat, 
                      FUN = mean, na.rm = TRUE)
names(expected)[2] <- "mean"
expected$OE <- "Exp"
observed <- aggregate(action_found ~ formal_in, 
                      data = dat, 
                      FUN = mean, na.rm = TRUE)
names(observed)[2] <- "mean"
observed$OE <- "Obs"
overall <- rbind(expected, observed)
overall <- select(overall, OE, formal_in, mean)
names(overall) <- c("Obs/Exp", "Formal/Informal", "Mean")
kable(overall, digits = 3)
```

Next, we break the categories down to identify which classes changed between
the expected and observed stages, e.g., if "maybe"s were, in reality, not 
observable:

```{r p2}
values <- c("No", "Maybe", "Yes")
for_exp <- table(filter(dat, formal_in == "formal")$reconcile)
for_obs <- table(filter(dat, formal_in == "formal")$action_found)
inf_exp <- table(filter(dat, formal_in == "informal")$reconcile)
inf_obs <- table(filter(dat, formal_in == "informal")$action_found)
new <- data_frame(
  OE = c(rep("expected", 6), rep("observed", 6)),
  `in/formal` = c(rep("formal", 3), rep("informal", 3),
           rep("formal", 3), rep("informal", 3)),
  observed = rep(values, 4),
  frequency = c(for_exp, inf_exp, for_obs, inf_obs)
)

ggplot(data = new, aes(x = observed, y = frequency)) +
  geom_bar(stat = "identity") +
  labs(x = "Action observability",
       y = "# consultations") +
  facet_grid(OE ~ `in/formal`) +
  theme_hc()
```

The "Maybe" categories for expected were mostly "No" when observed, for both 
formal and informal consultations.

```{r p3}
new
```

## Observability by Work Category 

### High variation among work categories

Determining the types of work that are most amenable to monitoring from the
standpoint of observability - whether the effects of the action are observable 
in aerial or satellite imagery - is a key challenge. Before showing 
observability rate, here is the sample size by work type and by formal/informal 
consultation:

```{r tmp}
n_obs <- as_data_frame(table(dat$work_category, dat$formal_in))
names(n_obs) <- c("work_cat", "formal_in", "n")

marg <- ggplot(n_obs, aes(x = n, y = work_cat)) +
  geom_lollipop(horizontal = TRUE, point.size = 3) +
  facet_grid(. ~ formal_in) +
  labs(x = NULL, y = "Sample size") +
  theme_hc()
marg

# Remove the y-axis labels to make compound fig and save for the ms.
# marg <- marge +
#   theme(axis.title.y=element_blank(),
#         axis.text.y=element_blank(),
#         axis.ticks.y=element_blank())
# ggsave("inst/figs/work_cat_counts.pdf", height = 9.75, width = 3.25)
```

And now the mean observability by work category and formal/informal consultation:

```{r obs_heatmap}
type_obs <- aggregate(
  action_found ~ work_category + formal_in,
  data = dat,
  FUN = mean, na.rm = TRUE
)
names(type_obs) <- c("category", "in/formal", "observability")

fig <- ggplot(type_obs, aes(y = category, x = `in/formal`, fill = observability)) +
  geom_tile(color = "white", size = 0.1) +
  labs(x = "Consultation Type", y = "") +
  scale_y_discrete(limits = levels(type_obs$category)) +
  scale_fill_viridis(name="Mean Observability") +
  theme_hc()
# fig
# ggsave("inst/figs/work_cat_obs.pdf", height = 9.75, width = 6.5)

# Use ggplotly for this HTML vignette:
ggplotly(fig)
```

There has been a slight decrease in observability over time, but that
decrease is not statistically significant:

```{r obs_time}
cor.test(dat$action_found, dat$FY)
```

## Expected Volume of Consultations

### We expect ~40% overall observability if all coordinates recorded...

```{r expected_overall}
get_observabilities <- function(dat, formal_in) {
  type_mean <- tapply(dat$action_found,
                      INDEX = dat$work_type,
                      FUN = mean, na.rm = TRUE)
  type_median <- tapply(dat$action_found,
                      INDEX = dat$work_type,
                      FUN = median, na.rm = TRUE)
  if (formal_in == "formal") {
      type_count <- tapply(dat$N_formal,
                           INDEX = dat$work_type,
                           FUN = mean, na.rm = TRUE)
  } else {
      type_count <- tapply(dat$N_consultations,
                           INDEX = dat$work_type,
                           FUN = mean, na.rm = TRUE)
  }
  expect_to_see_all <- type_mean * type_count

  # And these are the results at the highest level:
  tot_num_formal <- sum(type_count, na.rm = TRUE)
  exp_num_see <- sum(expect_to_see_all, na.rm = TRUE)
  obs_rate <- round(exp_num_see / tot_num_formal, 2)
  in_set <- round(tot_num_formal, 0)
  exp_obs <- round(exp_num_see, 0)
  return(c(obs_rate, in_set, exp_obs))
}

rows <- c("Observability | work type", 
          "# consultations in set",
          "# consultations expect to see effects")
exp_df <- data_frame(
  value = rows,
  formal = get_observabilities(formal, "formal"),
  informal = get_observabilities(informal, "informal")
)
exp_df
```

_Note:_ The number of consultations in each set is the number in the work 
categories we examined, not the total number of consultations with coordinates. 
There is a long tail of work categories with few consultations.

### ...but work type matters a lot

The "lollipops" indicate the number of consultations we expect to find for each
work category given the observability of work types and their frequency in 
consultations. The gray circles are the number of actions of one or more work
_types_ per work _category_, and the red vertical lines are the mean number 
of actions we expect to find for each work type within work category:

```{r availables, warning=FALSE, error=FALSE, message=FALSE}
make_scatter_df <- function(dat, cons_dat, formal_in) {
  type_mean <- tapply(dat$action_found,
                      INDEX = dat$work_type,
                      FUN = mean, na.rm = TRUE)
  type_median <- tapply(dat$action_found,
                      INDEX = dat$work_type,
                      FUN = median, na.rm = TRUE)
  if (formal_in == "formal") {
      type_count <- tapply(dat$N_formal,
                           INDEX = dat$work_type,
                           FUN = mean, na.rm = TRUE)
  } else {
      type_count <- tapply(dat$N_consultations,
                           INDEX = dat$work_type,
                           FUN = mean, na.rm = TRUE)
  }
  expect_to_see_all <- type_mean * type_count

  tmp_dat <- data.frame(type_count = as.vector(type_count),
                        type_mean = as.vector(type_mean),
                        work = names(type_count))
  work_cat_type <- data.frame(cat = cons_dat$work_category,
                              work = as.character(cons_dat$work_type))
  work_cat_type$uniq <- duplicated(work_cat_type$work)
  work_cat_type <- work_cat_type[work_cat_type$uniq == FALSE, ]
  tmp_dat <- inner_join(tmp_dat, work_cat_type, by = "work")
  return(tmp_dat)
}
form_obs_dat <- make_scatter_df(formal, form_cons, "formal")
form_obs_dat$expected <- form_obs_dat$type_count * form_obs_dat$type_mean
cat_max <- aggregate(expected ~ cat, 
                     data = form_obs_dat, 
                     FUN = max, na.rm = TRUE)
names(cat_max)[2] <- "cat_max"
cat_mean <- aggregate(expected ~ cat, 
                     data = form_obs_dat, 
                     FUN = mean, na.rm = TRUE)
names(cat_mean)[2] <- "cat_mean"
cat_sum <- aggregate(expected ~ cat, 
                     data = form_obs_dat, 
                     FUN = sum, na.rm = TRUE)
names(cat_sum)[2] <- "cat_sum"
form_obs_dat <- left_join(form_obs_dat, cat_max, by = "cat")
form_obs_dat <- left_join(form_obs_dat, cat_mean, by = "cat")
form_obs_dat <- left_join(form_obs_dat, cat_sum, by = "cat")
form_obs_dat$cat_ord <- reorder(form_obs_dat$cat, form_obs_dat$cat_sum)
form_obs_dat$in_formal <- "Formal"

########################
# The following includes code for the scatterplot of number of actions
# vs. observability. It is superceded by a lollipop chart in the informal
# consultation section below...
#
# tags$div(
#   plot_ly(data = form_obs_dat,
#           type = "scatter",
#           mode = "markers",
#           y = form_obs_dat$type_count,
#           x = round(form_obs_dat$type_mean, 2),
#           text = paste("Work type:", 
#                        form_obs_dat$work, 
#                        "<br>Work category:", 
#                        form_obs_dat$cat),
#           marker = list(
#             color = substr(
#               viridis(n = length(unique(form_obs_dat$cat))), 0, 7
#             ),
#             opacity = 0.6,
#             size = 20)
#           ) %>%
#   layout(yaxis = list(title = "# actions"),
#          xaxis = list(title = "Prop. observed"))
# )
```

```{r informal_consult_exp_type, warn=FALSE, error=FALSE, message=FALSE}
inform_obs_dat <- suppressWarnings(
  make_scatter_df(informal, inform_cons, "informal")
)

inform_obs_dat$expected <- inform_obs_dat$type_count * inform_obs_dat$type_mean
cat_max <- aggregate(expected ~ cat, 
                     data = inform_obs_dat, 
                     FUN = max, na.rm = TRUE)
names(cat_max)[2] <- "cat_max"
cat_mean <- aggregate(expected ~ cat, 
                     data = inform_obs_dat, 
                     FUN = mean, na.rm = TRUE)
names(cat_mean)[2] <- "cat_mean"
cat_sum <- aggregate(expected ~ cat, 
                     data = inform_obs_dat, 
                     FUN = sum, na.rm = TRUE)
names(cat_sum)[2] <- "cat_sum"
inform_obs_dat <- left_join(inform_obs_dat, cat_max, by = "cat")
inform_obs_dat <- left_join(inform_obs_dat, cat_mean, by = "cat")
inform_obs_dat <- left_join(inform_obs_dat, cat_sum, by = "cat")
inform_obs_dat$cat_ord <- reorder(inform_obs_dat$cat, inform_obs_dat$cat_sum)
inform_obs_dat$in_formal <- "Informal"

allocation <- rbind(form_obs_dat, inform_obs_dat)
cat_mean_rate <- aggregate(type_mean ~ cat, 
                     data = allocation, 
                     FUN = mean, na.rm = TRUE)
names(cat_mean_rate)[2] <- "cat_mean_rate"
allocation <- left_join(allocation, cat_mean_rate, by = "cat")
allocation$combo_ord <- reorder(allocation$cat, allocation$cat_sum)

marg <- ggplot(allocation, aes(x = cat_sum, y = combo_ord)) +
  geom_lollipop(horizontal = TRUE, point.size = 3) +
  geom_jitter(aes(x = type_count, y = combo_ord), 
             size = 3,
             alpha = 0.2, height = 0.1, width = 0) +
  geom_point(aes(x = cat_mean, y = combo_ord), 
             shape = "|", 
             size = 4,
             colour = "red") +
  facet_grid(. ~ in_formal, scales = "free") +
  labs(x = "No. Actions", y = NULL) +
  theme_hc()
marg

# tags$div(
#   plot_ly(data = inform_obs_dat,
#           type = "scatter",
#           mode = "markers",
#           y = inform_obs_dat$type_count,
#           x = round(inform_obs_dat$type_mean, 2),
#           text = paste("Work type:", 
#                        inform_obs_dat$work, 
#                        "<br>Work category:", 
#                        inform_obs_dat$cat),
#           marker = list(
#             color = substr(
#               viridis(n = length(unique(inform_obs_dat$cat))), 0, 7
#             ),
#             opacity = 0.6,
#             size = 20)
#           ) %>%
#   layout(yaxis = list(title = "# actions"),
#          xaxis = list(title = "Prop. observed"))
# )

```

## Types of Change

### Natural habitat to development is dominant type of change

We recorded the type of habitat change observed at each found action site to
evaluate the types of change that are consulted on:

```{r habitat_chg_type}
trans_hab <- function(x) {
    if (is.na(x) | is.null(x)) NA
    else if (x == 1) "natural -> natural"
    else if (x == 2) "natural -> agriculture"
    else if (x == 3) "natural -> development"
    else if (x == 4) "agriculture -> development"
    else "development -> development"
}

tmp <- sapply(dat$hab_chg, trans_hab)
conv_type <- table(tmp, dat$formal_in)
conv_type <- data_frame(
  conversion = row.names(conv_type),
  formal = conv_type[, 1],
  informal = conv_type[, 2]
)
conv_type
```

## Imagery Dates

### Some images are old, but most are recent

Knowing the distribution of the oldest average aerial images is useful for 
anticipating how far we can "turn the clock back" to track habitat changes at
consultation sites.

```{r img_date, warning=FALSE, error=FALSE, message=FALSE}
summary(dat$earliest_date, na.rm = T)

ggplot(dat, aes(earliest_date)) +
    geom_histogram(colour="white") +
    labs(x = "Earliest Aerial Image Date") +
    theme_hc()
```

